---
title: "ProjectReport"
output:
  html_document: default
  pdf_document: default
date: "2023-11-15"
---
## Introduction
In a state as large as Texas, it is no surprise that the use of an automobile is sometimes not only a luxury, but a necessity. With  a population as large as Texas, this means milliona of cars are on the roads daily and with this, comes an increase of accidents. This project is aiming to answer the following questions: Where do a lot of the accidents in Texas take place? Where is the area in Texas with more severe accidents? What time of day/year do most accidents occur? 

## Loading up our libraries. 
tidtverse I am  using to manipulate the data. maps I am using to map
```{r}

library(tidyverse)
library(maps)
library(ggplot2)

```

#Reading in our data set. 
This dataset contains accident information from February 2016 - March 2023. Link to dataset: [https://www.kaggle.com/datasets/sobhanmoosavi/us-accidents]. The original dataset is HUGE so I decided to filter it to only show accidents within the state of Texas.
```{r}
all_accident_data <- read_csv('US_Accidents_March23 3.csv')
tx_only_data <- filter(all_accident_data, State == "TX")


```
##Cleaning and formatting dataset
Let's try and make sure this data is clean and able to be worked with. I also converted the 'Start_Time' and 'End_Time' columns to date time format as to be easier to manipulate. The format of Start_Time and End_Time is in the following format: YYYY-dd-mm and is followed by the time stamp HH:mm:ss. I also extracted the Year from the values in this column.

```{r}
glimpse(tx_only_data)

tx_only_data$Start_Time <- as.POSIXct(tx_only_data$Start_Time, format="%Y-%m-%d %H:%M:%S")
tx_only_data$End_Time <- as.POSIXct(tx_only_data$End_Time, format="%Y-%m-%d %H:%M:%S")

tx_only_data$Year <- format(tx_only_data$Start_Time, "%Y")
```

So now that we have formatted this data, my first question was - How many accidents occurred in Texas year by year according to this dataset? I first grouped the accident data by year and arranged the values in descending order. 



```{r}
accidents_by_year_ranked <- tx_only_data %>% 
  group_by(Year) %>% 
  summarise(Accidents = n()) %>% 
  arrange(desc(Accidents))
accidents_by_year_ranked

```
As we can see, 2021 had the highest numbers of accidents reported in the state of Texas. 

Let's plot this data so we can visualize this a little better. 



```{r}
ggplot(accidents_by_year_ranked, aes(x = Year, y = Accidents, fill = Year)) + 
  geom_bar(stat = "identity") +
  xlab("Year") + 
  ylab("Accidents") + 
  ggtitle("Accidents in Texas by year")
```
## Projecting number of accidents for 2023

After looking at this chart,  I  remembered that the recording of data began in February of 2016 and ends in March of 2023. Since the recording of this data stopped in March of 2023, I thought it would be interesting to project the amount of accidents for the rest of 2023. 

First, I decided to calculate the average rate of accidents per year. To get a more accurate average, I  decided to remove the years that did not include data for the entire years which in this case are years 2016 and 2023.
 
```{r}
not_including_partial_years <- accidents_by_year_ranked %>% 
  filter(!Year %in% c("2016", "2023")) #removing 2016 and 2023 years 
not_including_partial_years
```

Started by adding the values in 'Accidents' column for years 2017-2022 and getting the amount of rows which would be the years we have in this dataset. 
```{r}
total_accidents_full <- sum(not_including_partial_years$Accidents) #adding  accidents from 2017-2022
number_of_years <- nrow(not_including_partial_years)
```
Then dividing the sum of the total accidents by the amount of years.
```{r}
yearly_average <- total_accidents_full/number_of_years #total number of accidents from 2017-2023 divided by the number of years
yearly_average

```
Using this calculation we can see that the annual average of reported accidents in the state of Texas is 86540.67. 

Now to calculate the projection, I started by calculating the monthly rate of accidents for the years that we have complete information for. 
```{r}
monthly_rate <- not_including_partial_years %>% #years 2017-2022
  mutate(MonthlyAccidents = Accidents / 12) %>%  #
  summarise(AverageMonthlyRate = mean(MonthlyAccidents)) #mean of monthly rate
monthly_rate
```
Then calculated the average monthly rate of accidents from 2017-2022
```{r}
average_monthly_rate <- monthly_rate$AverageMonthlyRate
average_monthly_rate
```
With this calculated monthly rate of accidents, I then calculated the estimated amount of accidents for the rest of 2023  and then added that number to the amount of accidents already reported until  March 2023 which yields us the result 78729. 
```{r}
accidents_until_march <- 13824 #amount of accidents reported until March  2023
estimated_accidents_rest_of_2023 <- average_monthly_rate * (12 - 3) #multiplying average monthly rate by amount of months remaining 

total_estimated_accidents_2023 <- accidents_until_march + estimated_accidents_rest_of_2023 #adding amount of accidents reported until march 2023 to the estimated amount for the rest of 2023
total_estimated_accidents_2023
```
I then created a new dataframe with the new projected accidents  for 2023 and plotted the results. 
```{r}
accidents_with_2023_projection <- rbind(not_including_partial_years, data.frame(Year = "2023", Accidents = total_estimated_accidents_2023))

ggplot(accidents_with_2023_projection, aes(x = Year, y = Accidents, group = 1)) +
  geom_line() +
  geom_point() +
  labs(title = "Projected Total Accidents for 2023", x = "Year", y = "Number of Accidents")
```
## Accidents by County
I decided to try and break down this data to observe which counties have a higher density of accidents within those counties.

 
```{r}
accidents_by_county <- tx_only_data %>% 
  
  group_by(County, Severity) %>% 
  summarise(Accidents= n(), .groups = 'drop')  
 


total_accidents_by_county <- accidents_by_county %>% 
  group_by(County) %>% 
  summarise(TotalAccidents = sum(Accidents)) %>% 
  arrange(desc(TotalAccidents))
total_accidents_by_county
```
Since there are a total of 254 counties within this massive state, I decided to focus on the top 10 counties with the most accidents.
```{r}
top_10_counties <- head(total_accidents_by_county, 10) 
top_10_counties
 
```
Let's plot the 10 counties with the highest amount of reported accidents. 
```{r}
ggplot(top_10_counties, aes(x = reorder(County, -TotalAccidents), y = TotalAccidents)) +
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(title = "Top 10 Counties by Total Accidents",
       x = "County", 
       y = "Total Accidents")
```
Just because I was curious, I decided to also find out the 10 counties with the least amount of accidents reported. 
```{r}
bottom_10_counties <- tail(total_accidents_by_county, 10) 
bottom_10_counties
 
bottom_10_counties_with_severity <- merge(bottom_10_counties, accidents_by_county, by = 'County')
bottom_10_counties_with_severity
```
Which will now give us a much less interesting bar chart.
```{r}
ggplot(bottom_10_counties, aes(x = reorder(County, -TotalAccidents), y = TotalAccidents)) +
  geom_bar(stat = "identity", fill = "steelblue") + 
  labs(title = "Bottom 10 Counties by Total Accidents",
       x = "County", 
       y = "Total Accidents")
```
Let's take a look at the  maps_data() method to see how we can create a heatmap of accidents by county. This method contains the name of the county under the subregion column and longitude and latitude information for that county.
```{r}
print(head(map_data("county", region = "texas"), 10))

```
Using the maps package, I quickly found out that the County data was provided in all lowercase while our dataset had the first letter in the county capitalized so first, I had to change the county names to all lowercase. 
```{r}
#library(maps)

tx_only_data_county_to_lower <- tx_only_data %>%
                     mutate(County = tolower(County))
tx_only_data_county_to_lower
```
Once the data was formatted correctly, I began by creating an outline of the state of Texas by using the map_data method that collects coordinates of Texas's borders

```{r}
tx_state_map <- map_data("state", region = "texas") #outline of the state of Texas

```
I grouped the data by county and also pulled in coordinate data of county boundaries which I then merged into a data frame called 'merged_data' in order to match this accident data to actual county boundary data. 

```{r}
accidents_by_county <- tx_only_data_county_to_lower %>% 
  
  group_by(County) %>% 
  summarise(Accidents= n(), mean_long = mean(Start_Lng), mean_lat = mean(Start_Lat)) %>% 
  arrange(desc(Accidents))


texas_counties_map <- map_data("county", region="texas")


merged_data <- merge(texas_counties_map, accidents_by_county, by.x = "subregion", by.y = "County")
```




And the end result:

```{r}
ggplot(merged_data, aes(x = long, y = lat, group = group, fill = Accidents)) +
    geom_polygon(color = "white") +
    geom_polygon(data = tx_state_map, aes(x = long, y = lat, group = group), 
                 fill = NA, color = "black", size = 0.5) +
   
    coord_fixed(0.8) +
    scale_fill_gradient(low = "green", high = "red") +
    labs(title = "Heat Map of Accidents by County in Texas", fill = "Number of Accidents") +
    theme_minimal()

```
 Now we can see that Harris County had the most total accidents in this data but let's see which counties have the highest severity of accidents per this data.
 Severity is defined in this dataset as the amount of time the accident had an effect on traffic. Essentially, the higher the number of severity, the more affect this accident had on traffic. 
```{r}
accidents_by_county_with_severity <- tx_only_data_county_to_lower %>% 
  
  group_by(County, Severity) %>% 
  summarise(Accidents= n(), .groups = 'drop')

accidents_by_county_with_severity

total_accidents_by_county_with_severity <- accidents_by_county_with_severity %>% 
  group_by(County) %>% 
  summarise(TotalAccidents = sum(Accidents)) %>% 
  arrange(desc(TotalAccidents))

top_10_again <- head(total_accidents_by_county_with_severity, 10)

top_10_counties_with_severity <- merge(top_10_again, accidents_by_county_with_severity, by = "County")
top_10_counties_with_severity

```

```{r}
ggplot(top_10_counties_with_severity, aes(x = reorder(County, -Accidents), y = Accidents, fill = as.factor(Severity))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~Severity, scales = "free_y") +
  scale_fill_brewer(palette = "Spectral") +
  theme_minimal() +
  labs(title = "Total Accidents by Severity Level in Top 10 Counties",
       x = "County",
       y = "Number of Accidents",
       fill = "Severity Level") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
After a little digging you can see that even though Harris county has more total accidents, Dallas county has disproportianetly more accidents with higher severity. 


## Accidents by City

After analyzing the data from a county view, I wanted to dig a little more and see the accident data by city. For simplicity, I decided to only look at the top 10 cities with the most accidents. 
```{r}
top_10_cities <- tx_only_data %>% 
  group_by(City) %>% 
  summarise(Accidents= n(), mean_long = mean(Start_Lng), mean_lat = mean(Start_Lat)) %>% 
  arrange(desc(Accidents)) %>% 
  head(5)

top_10_cities <- top_10_cities %>% 
  filter(!is.na(mean_long) & !is.na(mean_lat) & !is.na(Accidents))
top_10_cities
```

```{r}
tx_state_map <- map_data("state", region = "texas")

ggplot(data = tx_state_map) +
  geom_polygon(aes(x = long, y = lat, group = group), fill = "white", color = "black") +
  geom_point(data = top_10_cities, aes(x = mean_long, y = mean_lat, size = Accidents), color = "blue", alpha = 0.5) +
  geom_label(data = top_10_cities, aes(x = mean_long, y = mean_lat, label = City), vjust = -1, color = "blue") +
  scale_size_continuous(range = c(1, 10)) +  # Adjust the range for size scaling as needed
  labs(title = "Accidents by City in Texas", size = "Number of Accidents") +
  theme_minimal() +
  coord_fixed(1.3)


```

```{r}
bottom_10_cities <- tx_only_data %>% 
  group_by(City) %>% 
  summarise(Accidents= n(), mean_long = mean(Start_Lng), mean_lat = mean(Start_Lat)) %>% 
  arrange(desc(Accidents)) %>% 
  tail(10)
  

bottom_10_cities <- bottom_10_cities %>% 
  filter(!is.na(mean_long) & !is.na(mean_lat) & !is.na(Accidents))
bottom_10_cities


tx_state_map <- map_data("state", region = "texas")

ggplot(data = tx_state_map) +
  geom_polygon(aes(x = long, y = lat, group = group), fill = "white", color = "black") +
  geom_point(data = bottom_10_cities, aes(x = mean_long, y = mean_lat, size = Accidents), color = "blue", alpha = 0.4) +
  geom_label(data = bottom_10_cities, aes(x = mean_long, y = mean_lat, label = City), vjust = -1, color = "blue") +
  scale_size_continuous(range = c(1, 11)) +  # Adjust the range for size scaling as needed
  labs(title = "Accidents by City in Texas", size = "Number of Accidents") +
  theme_minimal() +
  coord_fixed(1.3)


```



 








## Are there any roads that are more prone to accidents than others?

By grouping the street names together, it provided me with streets that have the most accidents reported in the state. However, plotting this would be very difficult and less general than what I wanted so I decided to instread aggregate the Start_Lng and Start_Lat to get a more generalized total for accidents. 

```{r}
top_streets <- tx_only_data %>%
  group_by(Street, Start_Lat, Start_Lng) %>%
  summarise(Accidents = n(), .groups  = 'drop')  %>%
  arrange(desc(Accidents))
  
top_streets
```
After aggregating this data, it provided me with some more general information that contained mostly major highways. 

```{r}
accidents_agg <- top_streets %>%
  group_by(Street) %>%
  summarise(Accidents= n(), mean_long = mean(Start_Lng), mean_lat = mean(Start_Lat), .groups = 'drop') %>% 
  arrange(desc(Accidents)) 
  
accidents_agg

top_accidents_agg <- accidents_agg %>% 
  head(10)
top_accidents_agg
```


  
```{r}
ggplot(tx_state_map) +
  geom_polygon(aes(x = long, y = lat, group = group), fill = "white", color = "black") +
  geom_point(data = accidents_agg, aes(x = mean_long, y = mean_lat, color = Accidents), color = "blue", alpha = 0.4) +
  
  geom_label(data = top_accidents_agg, aes(x = mean_long, y = mean_lat, label = Street), vjust = -1, color = "blue") +
  labs(title = "Heatmap of Accidents on Texas Streets", x = "Longitude", y = "Latitude") +
  coord_fixed(1.3) + 
  scale_size_continuous(range = c(1, 10))

```
Unsurprisingly, major  highways in the state account for the majority of the accidents that occur. 

# Weather

To be able to determine if there is any relationship between weather conditions and a higher risk of accidents, I grouped all of the available weather conditions information by grouping the Weather_Condition column together, omitting any NA values and then ranking them in descending order. 

```{r}
weather_accidents <- tx_only_data %>% 
  group_by(Weather_Condition, Severity) %>% 
  na.omit() %>% 
  summarise(Accidents = n()) %>% 
  arrange(desc(Accidents)) %>% 
  head(5)

weather_accidents

ggplot(weather_accidents, aes(x = reorder(x = Weather_Condition, -Accidents ), y = Accidents, fill = Weather_Condition))+ 
  geom_bar(stat = "identity") +
  xlab("Weather Condition") + 
  ylab("Accidents") + 
  ggtitle("Accidents in Texas by weather condition")
```
By this plot, we can see that the vast majority of accidents occur during 'Fair' weather conditions.

I also wanted to observe if there was any relationship between the weather condition and the severity of the accident. 
```{r}
accidents_by_weather <- tx_only_data %>% 
  
  group_by(Weather_Condition, Severity) %>% 
  na.omit() %>% 
  summarise(Accidents= n(), .groups = 'drop')  
 
accidents_by_weather

total_accidents_by_weather <- accidents_by_weather %>% 
  group_by(Weather_Condition) %>% 
  summarise(TotalAccidents = sum(Accidents)) %>% 
  arrange(desc(TotalAccidents))



top_10_weather_accidents <- head(total_accidents_by_weather, 10) 
top_10_weather_accidents
 
top_10_weather_accidents_with_severity <- merge(top_10_weather_accidents, accidents_by_weather, by = 'Weather_Condition')
top_10_weather_accidents_with_severity

ggplot(top_10_weather_accidents_with_severity, aes(x = reorder(Weather_Condition, -Accidents), y = Accidents, fill = as.factor(Severity))) +
  geom_bar(stat = "identity", position = position_dodge()) +
  facet_wrap(~Severity, scales = "free_y") +
  scale_fill_brewer(palette = "Spectral") +
  theme_minimal() +
  labs(title = "Total Accidents by Severity Level by top 10 Weather Conditions",
       x = "Weather Condition",
       y = "Number of Accidents",
       fill = "Severity Level") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
And here we can see that broken down by weather conditon and severity, there is not any obseravble relatioship between weather condition and severity of the accident. 

## Hour of the day

Accidents can occur any time a car is on the road but is there any specific hour(s) that have a higher prevalence of accidents that others?
In order to find this out, I had to format our data once again by extracting the hour from the Start_Time values. Once those hours were extracted, I was then able to group them all and summarise the hourly accidents. 

```{r}
tx_only_data$Hour <- format(tx_only_data$Start_Time, "%H")

hourly_accidents <- tx_only_data %>% 
  group_by(Hour) %>% 
  summarise(Accidents = n())
hourly_accidents
```
Once I was able to get the total amount of accidents per hour, I then plotted this information. 
```{r}
ggplot(hourly_accidents, aes(x = Hour, y = Accidents)) +
  geom_bar(stat = "identity") +
  xlab("Hour of Day") +
  ylab("Number of Accidents") +
  ggtitle("Accidents Distribution by Hour of Day")

```
Now we can clearly see that the hours of 8am and 5pm have the highest number of accidents recorded. This is more than likely due to more people being on the road during 'rush' hours which is the most common hours that people are traveling to and from work. 

## Monthly data

Lastly, I wanted to anlyze if there were any months throughout the year that had more reported accidents than others. 

Similiarly to the hourly analysis, I had to extract the Month from our Start_Time column and group each month and tally up the amount of accidents per month. 

```{r}
tx_only_data$Month <- format(tx_only_data$Start_Time, "%m")
monthly_accidents <- tx_only_data %>% 
  group_by(Month) %>% 
  summarise(Accidents = n()) %>% 
  arrange(desc(Accidents))
monthly_accidents
```

```{r}
ggplot(monthly_accidents, aes(x = Month, y = Accidents, fill = Month)) +
  geom_bar(stat = "identity") +
  
  xlab("Month") +
  ylab("Number of Accidents") +
  ggtitle("Accidents Distribution by Month of the Year") +
  scale_x_discrete(labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec'))
```

With our table and plot, we can see that the top 3 months in which accidents have occured have been in November, December, and January.

## Summary

In the end, there is no perfect predictor of when an accident will occur but in terms of observing the Lone Star State a few things are clear. Accidents in 2023 are predicted to be lower on average than previous years. The top 3 cities with most accidents reported (Houston, Dallas, and Austin) account for 93%, 84%, and 94% of their respective counties (Harris, Dallas, Travis) accidents. Times in which majority of people are traveling to and from to work (8am & 5pm) are the hours of the day in which more accidents are reported. And lastly, that the months November, December, and January which are also the holiday season months see the most accidents happen within the state. 








